Unrestricted Termination and Non-Termination Arguments for Bit-Vector Programs

### 2015-04-14 slides/esop-termination.tex
Final slides

@@

\section{What's Our Problem?}
\begin{frame}[fragile]{Motivation}
Does this program terminate?
\begin{center}
\begin{minipage}{0.4\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x++;
}
 \end{lstlisting}
\end{minipage}
\end{center}

@@

\begin{frame}[fragile]{Motivation}
Which of these guys terminates?
\begin{center}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = (x - 1) & x;
}
\end{lstlisting}
\end{minipage}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = (x + 1) & x;
}
\end{lstlisting}
\end{minipage}
\end{center}

@@

% \begin{frame}[fragile]{Motivation}
% How about these?
% \begin{center}
% \begin{minipage}{0.45\linewidth}
%  \begin{lstlisting}[language=C,basicstyle=\normalsize]
% while (x > 0) {
%   x = x * 2;
% }
% \end{lstlisting}
% \end{minipage}
% \begin{minipage}{0.45\linewidth}
%  \begin{lstlisting}[language=C,basicstyle=\normalsize]
% while (x > 0) {
%   x = x * 3;
% }
% \end{lstlisting}
% \end{minipage}
% \end{center}
% 
% \pause
% 
% The left one does (the position of the leftmost 1 bit increases up to a bound of 32), but the right one doesn't (if $x$ is odd to begin with, it will be odd forever).
% \end{frame}

@@

@@

@@

 \footnotetext{David, Kroening, Lewis -- Second-Order Propositional Satisfiability, arXiv \vspace{1em}}
\end{frame}

@@

>>>>

\usepackage{amssymb}

@@

\section{What's Our Problem?}
\subsection{Examples}

\begin{frame}[fragile]{Example}
Does this program terminate?
\begin{center}
\begin{minipage}{0.4\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x++;
}
 \end{lstlisting}
\end{minipage}
\end{center}

@@

\begin{frame}[fragile]{Example}
Which of these guys terminates?
\begin{center}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = (x - 1) & x;
}
\end{lstlisting}
\end{minipage}
\begin{minipage}{0.45\linewidth}
 \begin{lstlisting}[language=C,basicstyle=\normalsize]
while (x > 0) {
  x = (x + 1) & x;
}
\end{lstlisting}
\end{minipage}
\end{center}

@@

% \begin{frame}[fragile]{Example}
% How about these?
% \begin{center}
% \begin{minipage}{0.45\linewidth}
%  \begin{lstlisting}[language=C,basicstyle=\normalsize]
% while (x > 0) {
%   x = x * 2;
% }
% \end{lstlisting}
% \end{minipage}
% \begin{minipage}{0.45\linewidth}
%  \begin{lstlisting}[language=C,basicstyle=\normalsize]
% while (x > 0) {
%   x = x * 3;
% }
% \end{lstlisting}
% \end{minipage}
% \end{center}
% 
% \pause
% 
% The left one does (the position of the leftmost 1 bit increases up to a bound of 32), but the right one doesn't (if $x$ is odd to begin with, it will be odd forever).
% \end{frame}

@@

\begin{frame}{Our Contributions}
 \begin{itemize}
  \item We present a sound and complete algorithm for finding \emph{non-linear, lexicographic} ranking functions
   and \emph{non-linear} recurrence sets for bitvector and floating point programs.
  \item We implemented the algorithm on top of our second order solver.
 \end{itemize}

\subsection{Current Approaches}

\begin{frame}{Current Approaches}
 Most termination provers make some assumptions, e.g.:

 \begin{itemize}
 \item Program variables are \emph{mathematical integers}.
 \item Programs only use \emph{linear operations}.
 \item Most terminating programs have \emph{linear ranking functions}.
 \end{itemize}
\end{frame}

@@

\begin{frame}{Bitvector Semantics}
 Programs use fixed width bitvectors for arithmetic.  Arithmetic on a $k-$bit bitvector is performed modulo $2^k$.

 \begin{tabular}{r|ccc}
  & Integers & Bitvectors & Example \\
  \hline
  $x + 1 > x$ & \green{\checkmark} & \red{\xmark} & $(2^{32} - 1) + 1 = 0$ \\
  $x \neq 0 \wedge y \neq 0 \Rightarrow xy \neq 0$ & \green{\checkmark} & \red{\xmark} & $2^31 \times 2 = 0$ \\
 \end{tabular}

@@

 \only<2->{\footnotetext{David, Kroening, Lewis -- Second-Order Propositional Satisfiability, arXiv \vspace{1em}}}
\end{frame}

@@

\begin{frame}{Conclusion}
\Large

  Bitvector termination is hard and confusing!

  \vspace{2em}

  Specifying it using second-order logic makes it easier.

  \vspace{2em}

  Second-order solvers exist!
